XRONOS: An Open Data Infrastructure for Archaeological Chronology

### 2025-06-09 paper.qmd
Split table 1 to separate file

In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of version 1.2.0.

@@

```{r tbl-c14-datasets}
#| tbl-cap: Summary of published compilations of radiocarbon dates. For full data, see supplementary materials.
# References in tables broken by:
# https://github.com/quarto-dev/quarto-cli/issues/9342
c14_datasets |>
  transmute(
    database = glue("[{name}]({url})"),
    publication_year, n_dates, citations
  ) |>
  arrange(publication_year) |>
  gt() |>
  cols_label(
    publication_year = "published",
    citations = "references",
    n_dates = "dates"
    ) |>
  cols_label_with(everything(), str_to_sentence) |>
  sub_missing(everything()) |>
  fmt_markdown(c(database, citations)) |>
  fmt_number(c(n_dates), decimals = 0) |>
  cols_width(
    database ~ pct(40),
    publication_year ~ pct(15),
    n_dates ~ pct(15),
    citations ~ pct(30)
  ) |>
  cols_align(
    "left", c(database)
  ) |>
  tab_options(
    table.font.size = 13
  )
```

@@

<!-- TODO: add references? -->

>>>>

In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of [version 1.1.0](https://github.com/xronos-ch/xronos.rails/releases/tag/v1.1.0).

@@

```{r tbl-c14-datasets}
#| tbl-cap: Summary of published compilations of radiocarbon dates. For full data, see supplementary materials.
# References in tables broken by:
# https://github.com/quarto-dev/quarto-cli/issues/9342
#c14_datasets |>
#  transmute(
#    database = glue("[{name}]({url})"),
#    publication_year, n_dates, citations
#  ) |>
#  arrange(publication_year) |>
#  gt() |>
#  cols_label(
#    publication_year = "published",
#    citations = "references",
#    n_dates = "dates"
#    ) |>
#  cols_label_with(everything(), str_to_sentence) |>
#  sub_missing(everything()) |>
#  fmt_markdown(c(database, citations)) |>
#  fmt_number(c(n_dates), decimals = 0) |>
#  cols_width(
#    database ~ pct(40),
#    publication_year ~ pct(15),
#    n_dates ~ pct(15),
#    citations ~ pct(30)
#  ) |>
#  cols_align(
#    "left", c(database)
#  ) |>
#  tab_options(
#    table.font.size = 13
#  )
"omitted from manuscript – see attached file `table1.html`"
```

@@

### 2025-06-09 paper.tex
Split table 1 to separate file

In the remainder of this paper, we describe the concept and
implementation of XRONOS in relation to the state of the art in open
chronometric data in archaeology, and evaluate our progress in achieving
these goals as of writing. Since we envisage both XRONOS as a dataset
and XRONOS as software to be continually developing resources, the
description here should be read as a `snapshot' of the project as of
version 1.2.0.

@@

\fontsize{9.8pt}{11.7pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.40\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
Database & Published & Dates & References \\ 
\midrule\addlinespace[2.5pt]
\href{https://bda.huma-num.fr/}{Base de Données Archéologique} & 1994 & 7,000 & @Perrin2019 \\ 
\href{https://andesc14.pl}{ANDES 14C} & 1994 & 5,800 & @MichczynskiEtAl1995 \\ 
\href{http://www.tayproject.org/enghome.html}{Archaeological Settlements of Turkey} & 1998 & 1,600 & @TanindiErdogu2005 \\ 
\href{https://www.canadianarchaeology.ca/}{Canadian Archaeological Radiocarbon Database} & 1999 & 171,500 & @GajewskiEtAl2011; @KellyEtAl2022 \\ 
\href{https://radonb.ufg.uni-kiel.de/}{RADO.NB (incl. RADON and RADON-B)} & 1999 & 34,200 & @Raetzel-Fabian1999; @RADON; @RADONB; @RADO.NB \\ 
\href{https://www.waikato.ac.nz/nzcd}{New Zealand Radiocarbon Database} & 2000 & 2,000 & @McFadgenEtAl2000 \\ 
\href{http://web.archive.org/web/20080509082232/http://www.canew.org/index.html}{International Central Anatolian Neolithic e-Workshop databases} & 2001 & 1,000 & @ReingruberThissen2005; @ReingruberThissen2009 \\ 
\href{https://ees.kuleuven.be/geography/projects/14c-palaeolithic/}{Radiocarbon Palaeolithic Europe Database} & 2002 & 17,900 & @Vermeersch2024 \\ 
\href{https://zenodo.org/doi/10.5281/zenodo.7215741}{CalPal database} & 2002 & 49,800 & @Weninger2022 \\ 
\href{http://context-database.uni-koeln.de}{CONTEXT} & 2002 & 2,900 & @BohnerSchyle2004 \\ 
\href{http://pidba.utk.edu/dating.htm}{Paleoindian Database of the Americas} & 2003 & 1,300 & @AndersonEtAl2010 \\ 
\href{https://archaeologydataservice.ac.uk/archives/view/austarch_na_2014/}{AustArch (1, 2, and 3)} & 2008 & 5,000 & @WilliamsEtAl2008; @WilliamsSmith2012; @WilliamsSmith2013 \\ 
\href{https://sites.google.com/site/chapplearchaeology/irish-radiocarbon-dendrochronological-dates}{Irish Radiocarbon \& Dendrochronological Dates} & 2010 & 10,700 & @IRDD \\ 
\href{https://www.exoriente.org/associated_projects/ppnd.php}{Platform for Neolithic Radiocarbon Dates (PPND)} & 2010 & 800 & @Benz2010 \\ 
\href{http://www.paleoanthro.org/media/journal/content/PA20110001.pdf}{PACEA Geo-Referenced Radiocarbon Database} & 2011 & 6,000 & @dErricoEtAl2011 \\ 
\href{https://doi.org/10.1016/j.quaint.2012.08.2052}{Peru archaeological radiocarbon database} & 2013 & 300 & @RademakerEtAl2013 \\ 
\href{https://telearchaeology.org///EUBAR/}{EUBAR C14 database} & 2014 & 1,700 & @CapuzzoEtAl2014 \\ 
\href{https://doi.org/10.1016/j.quascirev.2014.05.015}{Wang 2014} & 2014 & 4,700 & @WangEtAl2014 \\ 
\href{http://www.14sea.org/}{14SEA} & 2015 & 3,000 & @ReingruberThissen2017 \\ 
\href{https://discovery.ucl.ac.uk/id/eprint/1469811/}{EUROEVOL} & 2015 & 14,100 & @ManningEtAl2016 \\ 
\href{https://doi.org/10.1016/j.quascirev.2015.06.022}{Flohr et al. 2015} & 2015 & 3,000 & @FlohrEtAl2016 \\ 
\href{https://doi.org/10.1016/j.quaint.2014.09.076}{South Central Andes Radiocarbon} & 2015 & 1,700 & @GayoEtAl2015 \\ 
\href{https://github.com/dirkseidensticker/aDRAC}{Archive des datations radiocarbones d'Afrique centrale} & 2016 & 1,900 & @SeidenstickerSchmid2021 \\ 
\href{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/NJLNRJ}{KITE East Africa} & 2016 & 800 & @CourtneyMustaphi2016 \\ 
\href{https://doi.org/10.14324/000.ds.1570274}{Palmisano et al. 2017} & 2017 & 1,900 & @PalmisanoEtAl2017 \\ 
\href{http://vmtropicar-proto.ird.fr/archeologie/}{Plateforme des datations archéologiques intertropicales} & 2017 & 1,300 & @deSaulieuEtAl2017 \\ 
\href{http://www.idearqueologia.org/c14/}{Archivo de Dataciones Radiocarbónicas de la Prehistoria Recient (IDEARQ)} & 2017 & — & @UriarteGonzalezEtAl2017 \\ 
\href{https://www.rekihaku.ac.jp/up-cgi/login.pl?p=param/esrd/db_param}{Database of Radiocarbon Dates Published in Japanese Archaeological Reports} & 2018 & 44,000 & @Kudo2018; KudoEtAl2023 \\ 
\href{https://c14.arch.ox.ac.uk/sadb/db}{Southern African Radiocarbon Database} & 2019 & 2,700 & @LoftusEtAl2019 \\ 
\href{https://doi.org/10.1016/j.quascirev.2019.105878}{Douglass et al. 2019} & 2019 & 200 & @DouglassEtAl2019 \\ 
\href{https://theia.arch.cam.ac.uk/MedAfriCarbon/}{MedAfriCarbon} & 2020 & 1,600 & @LucariniEtAl2020 \\ 
\href{https://github.com/clipo/rapanui-radiocarbon}{A repository of radiocarbon data for Rapa Nui} & 2020 & 800 & @Lipo2020 \\ 
\href{https://github.com/jgregoriods/rxpand}{rxpand} & 2020 & 2,800 & @GregoriodeSouza2020 \\ 
\href{https://core.tdar.org/project/455305/mesoamerican-radiocarbon-database-mesorad}{Mesoamerican Radiocarbon Database (MesoRAD)} & 2020 & 1,800 & @HoggarthEtAl2021 \\ 
\href{https://zenodo.org/records/4541470}{AgriChange} & 2020 & 3,700 & @Martinez-GrauEtAl2021 \\ 
\href{https://github.com/ercrema/jomonPhasesAndPopulation}{Crema \& Kobayashi 2020} & 2020 & 2,100 & @CremaEtAl2016 \\ 
\href{https://rdr.ucl.ac.uk/articles/dataset/Dataset_for_An_Aegean_history_and_archaeology_written_through_radiocarbon_dates/12489137/1}{Katsianis et al. 2020} & 2020 & 3,200 & @KatsianisEtAl2020 \\ 
\href{https://www.arqueodata.com/}{ArqueoData} & 2021 & 800 & @Alcantara2021 \\ 
\href{https://github.com/philriris/caribbean-14C}{caribbean-14C} & 2021 & 2,100 & @Riris2021 \\ 
\href{https://github.com/apalmisano82/NERD}{Near East Radiocarbon Dates (NERD)} & 2021 & 11,000 & @PalmisanoEtAl2022a \\ 
\href{https://doi.org/10.13131/archelogicadata-yb11-yb66}{NeoNet} & 2021 & 2,500 & @HuetEtAl2022; @HuetEtAl2024 \\ 
\href{https://github.com/ercrema/NeolithicKoreaDemography}{Kim et al. 2021} & 2021 & 900 & @KimEtAl2021 \\ 
\href{https://doi.org/10.1371/journal.pone.0251407}{Cochrane et al. 2021} & 2021 & 100 & @CochraneEtAl2021 \\ 
\href{https://www.waikato.ac.nz/research/research-services-facilities/radiocarbon-dating/research//nz-radiocarbon-database}{Aotearoa New Zealand Radiocarbon Databas} & 2022 & 4,100 & @PetcheyEtAl2022 \\ 
\href{https://github.com/apalmisano82/AIDA}{Archive of Italian Radiocarbon Dates (AIDA)} & 2022 & 4,000 & @PalmisanoEtAl2022 \\ 
\href{https://www.p3k14c.org/}{p3k14c} & 2022 & 179,700 & @BirdEtAl2022 \\ 
\href{https://research.jcu.edu.au/data/published/7c74f590a2ba11edb22c156e754c4bda/}{Torres Strait Radiocarbon Database} & 2022 & 300 & @LinnenluckeEtAl2023 \\ 
\href{http://holoceno.iaas.ull.es/14Canarias_ULL/}{14Canarias} & 2023 & 700 & @Pardo-GordoEtAl2023 \\ 
\href{https://doi.org/10.1016/j.jasrep.2023.103944}{Hoebe et al. 2023} & 2023 & 6,500 & @HoebeEtAl2023 \\ 
\href{https://core.tdar.org/collection/71234}{Bolivian Radiocarbon Database} & 2023 & 3,000 & @Capriles2023 \\ 
\href{https://doi.org/10.5281/zenodo.8334722}{An Annotated Compilation of Chronometric Dates for the Middle-Upper Palaeolithic Transition (45-30 ka BP) in Northern Iberia (Spain)} & 2023 & 200 & @Diaz-RodriguezEtAl2023 \\ 
\href{https://doi.org/10.1371/journal.pone.0291956}{Großman et al. 2023} & 2023 & 3,400 & @GrossmannEtAl2023 \\ 
\href{https://doi.org/10.1080/0067270X.2023.2215649}{Datations absolues, Inventaires archéologiques et Bibliographies en Afrique Centrale} & 2023 & 1,800 & @ClistEtAl2023 \\ 
\href{https://doi.org/10.1016/j.quaint.2024.01.012}{Updated Peru archaeological radiocarbon database} & 2024 & 500 & @Rademaker2024 \\ 
\href{https://banadora.mom.fr/}{Banque Nationale de Données Radiocarbone pour l'Europe et le Proche Orient} & — & — & @BANADORA \\ 
\href{https://c14.arch.ox.ac.uk/database}{Oxford Radicoarbon Accelerator Unit database} & — & 8,500 & @GillespieEtAl1984; @BronkRamseyEtAl2009 \\ 
\href{https://c14.arch.ox.ac.uk/egyptdb/db.php}{Egyptian Radiocarbon Database} & — & 1,600 & @RamseyEtAl2010; @DeeEtAl2012; @DeeEtAl2013 \\ 
\href{https://c14.arch.ox.ac.uk/database}{NERC Radiocarbon Facility} & — & 6,100 & @GarnettEtAl2023 \\ 
\href{http://www.adias-uae.com/radiocarbon.html}{Abu Dhabi Islands Archaeological Survey Radiocarbon Database} & — & 100 & @Al-Abyadh-Balghelam-Dalma-JebelEtAl \\ 
\href{http://c14.kikirpa.be/}{Royal Institute for Cultural Heritage web based Radiocarbon database} & — & 7,300 & @VanStrydonckDeRoock \\ 
\href{https://telearchaeology.org///c14/}{La base de dades radiocarbòniques de Catalunya} & — & — & @BarceloAlvarezEtAl2013 \\ 
\bottomrule
\end{tabular*}

@@

Looking beyond the near term, we have endeavoured to create a
sustainable infrastructure that can be maintained by a wider scholarly
commons -- though we must acknowledge that this is a difficult problem,
and one that is as much organisational than technical. The source code
for all the software components of the system are available online (at
\url{https://github.com/xronos-ch}) under open licenses. The databases
of the instance at \url{https://xronos.ch} are also archived with Zenodo
(\emph{link omitted for blind review}). With these two resources we
reduce XRONOS' `bus factor'; if we are not able to continue operating
XRONOS, somebody else can fully recreate it. Equally importantly, we
enable the `right to fork' should others wish to take the software
and/or database in another direction. But aside from these extreme
scenarios, the long-term sustainability of XRONOS is contingent on the
existence of a community of scholars that use and contribute back to it.

@@

@@

This work was funded by the Swiss National Science Foundation
(\href{https://data.snf.ch/grants/grant/198153}{SNSF Project \#198152})
and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz,
Project `Time and Temporality').

>>>>

In the remainder of this paper, we describe the concept and
implementation of XRONOS in relation to the state of the art in open
chronometric data in archaeology, and evaluate our progress in achieving
these goals as of writing. Since we envisage both XRONOS as a dataset
and XRONOS as software to be continually developing resources, the
description here should be read as a `snapshot' of the project as of
\href{https://github.com/xronos-ch/xronos.rails/releases/tag/v1.1.0}{version
1.1.0}.

@@

\begin{verbatim}
[1] "omitted from manuscript – see attached file `table1.html`"
\end{verbatim}

@@

Looking beyond the near term, we have endeavoured to create a
sustainable infrastructure that can be maintained by a wider scholarly
commons -- though we must acknowledge that this is a difficult problem,
and one that is as much organisational than technical. The source code
for all the software components of the system are available online (at
\url{https://github.com/xronos-ch}) under open licenses. The databases
of the instance at \url{https://xronos.ch} are also archived with Zenodo
at \url{https://doi.org/10.5281/zenodo.14282598}. With these two
resources we reduce XRONOS' `bus factor'; if we are not able to continue
operating XRONOS, somebody else can fully recreate it. Equally
importantly, we enable the `right to fork' should others wish to take
the software and/or database in another direction. But aside from these
extreme scenarios, the long-term sustainability of XRONOS is contingent
on the existence of a community of scholars that use and contribute back
to it.

@@

We are also grateful to the two anonymous reviewers of our manuscript,
whose comments have improved the finished paper.

@@

This work was funded by the Swiss National Science Foundation
(\href{https://data.snf.ch/grants/grant/198153}{SNSF Project \#198152};
\href{https://data.snf.ch/grants/grant/194326}{SNSF Project \#194326})
and the University of Bern (UniBE Initiator Grant 2019).
